---
title: "Robust PCA and image compression"
subtitle: "Section 7"
author: "Georgios Papadopoulos | "
date: "2025-11-16"
output: github_document
---

*Applying classical and robust PCA to handwriting features and compression based on PCA compression to false-color xray images*

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.path = "figures/"
  )
```

# 1. PCA of handwriting features

The darwinM dataset contains 19 handwriting measurements collected from two groups/classes of participants: healthy people `H` and patients diagnosed with Alzheimer’s `P`. Each participant completed 25 standardized writing tasks and our data has the median per measurement/variable. The dataset includes variables describing timing, speed, acceleration, pressure, jerk etc.

With following `summary(PCA)` I can conclude that cumulative proportion of PC1 and PC2 is 0.61, meaning that the first two components account for 61% of total variance in the data. So a large part of the multivariate dataset can be represented in a two-dimensional PCA plot, making PC1 and PC2 suitable for visualizing.

## 1.1 Classical PCA and group separation

I first produced the standard `biplot(pca)`. From the enlarged biplot it becomes clear which variables dominate PC1 and PC2:

-   The variables with the strongest loadings on PC1 are the ones that point in horizontal direction. pressure_mean, pressure_var, paper_time, air_time, disp_index, gmrt_in_air, gmrt_on_paper are the variables that most contribute to PC1

-   The variables with the strongest loadings on PC2 are the ones that point in vertical direction. mean_jerk_in_air, mean_jerk_on_paper, max_x_extension, max_y_extension, mean_speed_in_air\_ mean_speed_on_paper, mean_acc_in_air, mean_acc_on_paper are the variables that contribute to PC2

So simplified we can vaguely say that PC1 is responsible capturing pressure, writing time and gmrt. Whereas PC2 is responsible for maximal extension, jerk, speed and accelaration.

```{r}
load("darwinM.RData")

X <- darwinM[, setdiff(names(darwinM), "class")]

pca <- prcomp(X, center = TRUE, scale. = TRUE)

summary(pca)

biplot(pca, cex = 0.7)


```

To doublecheck the loadings I visualized them in barplots like at the last report.

```{r}
loadings <- pca$rotation[,1:2]

barplot(t(loadings),
        beside = TRUE,
        col = c("beige", "wheat"),
        names.arg = rownames(loadings),
        las = 2, cex.names = 0.5,
        main = "Loadings for PC1 and PC2")
legend("topright",
       legend = c("PC1", "PC2"),
       fill = c("beige", "wheat"))
```

Since the base R biplot does not color the groups, I decided to use the ggbiplot library so we can also see where are both classes of healthy people and patients with Alzheimer. This clarified how the two groups are positioned in the PCA space: healthy subjects lie mostly on the right side and patients on the left side

```{r}
library(ggbiplot)

ggbiplot(pca, groups = darwinM$class, ellipse = TRUE) +
  scale_color_manual(values = c("H" = "limegreen", "P" = "tomato")) +
  scale_fill_manual(values = c("H" = "limegreen", "P" = "tomato")) + theme_minimal()
```

## 1.2 Robust PCA diagnostics by group

I splitted the dataset by using the labels H and P. Then I removed the 19th variable which was the `class` variable. Afterwards applied robust PCA and created the diagnostic plots.

As mentioned on page 82/129: the cutoff for the Score Distance SD is based on the $\chi^2_{k,0.975}$ threshold, since SD behaves like a Mahalanobis distance in the PCA score space. For the Orthogonal Distance OD, a cutoff is derived from the median and MAD of $OD^{2/3}$ which is approximately normal. Observations that pass either cutoff are flagged as outliers. In our case it is for OD\>4 or SD\>2.7

-   Observations can appear as outliers in score distance when their projection onto the first two principal components lies far from the center of the data cloud. This means that, within the PCA subspace, these points have unusually large component scores compared to the rest of the group.

-   Observations can appear as outliers in orthogonal distance when that observation is poorly represented by the first two PCs, meaning much of its variation is not included in the PCA space (here PC1 and PC2). So these observations are captured in other PCs, like PC3 or PC4.

In the healthy group, observations 165, 136, 9, and 68 go over the orthogonal distance cutoff, which means they contain variation that is not captured by the first two PCs. Healthy observations 103 and 166 cross the score-distance cutoff, so they lie unusually far away within the PC1 PC2 space. No healthy subject crosses both cutoffs at the same time.

In the patient group, observation 4 crosses the score-distance cutoff, while observations 7 and 19 cross both the score-distance and orthogonal-distance cutoffs. These two patients therefore stand out the most, because they are extreme both within the PC1–PC2 space and outside of it.

```{r}
library(rrcov)

healthy_df <- subset(darwinM, class == "H")
patient_df <- subset(darwinM, class == "P")

healthy  <- as.matrix(subset(darwinM, class=="H")[ , -19])
patients <- as.matrix(subset(darwinM, class=="P")[ , -19])

rpca_healthy  <- PcaHubert(healthy,  k = 2, scale = TRUE)
rpca_patients <- PcaHubert(patients, k = 2, scale = TRUE)

par(mfrow = c(1, 2))

plot(rpca_healthy,  main = "Robust PCA for healthy")
plot(rpca_patients, main = "Robust PCA for patients")

```

# 2. PCA based compression of an X-ray image

By  loading the xray pnm file which contains the picture of the xray, I conclude from the structure that indeed it consists of three matrices inside named red, blue, green. We see that the structure of each color matrix has p = 746 variables such as v1, v2, ..., v746 and n = 1023 observations.

## 2.1 Reading and visualizing the image

```{r}
library(pixmap)
x <- read.pnm("xray.pnm")
plot(x)
str(x)

R <- x@red
G <- x@green
B <- x@blue
```

## 2.2 PCA reconstruction of RGB channels

I selected prcomp() because it performs PCA with SVD which is stable for large matrices. Scaling the data is not necessary because scaling would artificially equalize the variances of the pixel columns and distort the brightness and color structure of the image.

It is important to keep the original variance of the pixel intensities, because this variance has the visual information we want to keep in the picture reconstruction.

```{r}
pca_R <- prcomp(R, center = TRUE, scale. = FALSE)
pca_G <- prcomp(G, center = TRUE, scale. = FALSE)
pca_B <- prcomp(B, center = TRUE, scale. = FALSE)
```

To select k, I visuallized the cumulative variance of each color and for k=10 all first 10 PCs explain more than 90% of the variance in their matrix.

```{r}
par(mfrow = c(1, 3))

#red
var_exp_red <- pca_R$sdev^2 / sum(pca_R$sdev^2)
cum_var_red <- cumsum(var_exp_red)

plot(cum_var_red[1:746],
     xlab = "746 Principal Components",
     ylab = "Cumulative variance",
     main = "Cum. variance  - red")
abline(h = 0.9, col = "red", lty = 2)

#green
var_exp_green <- pca_G$sdev^2 / sum(pca_G$sdev^2)
cum_var_green <- cumsum(var_exp_green)

plot(cum_var_green[1:746],
     xlab = "746 Principal Components",
     ylab = "Cumulative variance",
     main = "Cum. variance  - green")
abline(h = 0.9, col = "red", lty = 2)

#blue
var_exp_blue <- pca_B$sdev^2 / sum(pca_B$sdev^2)
cum_var_blue <- cumsum(var_exp_blue)

plot(cum_var_blue[1:746],
     xlab = "746 Principal Components",
     ylab = "Cumulative variance",
     main = "Cum. variance - blue")
abline(h = 0.9, col = "red", lty = 2)

```

To reconstruct X with the first k PCs, I used following formulas (5.22) $Z=(X-1\bar{x}^\top)\,\hat{\Gamma}$

This is exactly what prcomp() computes:

-   `pca$x` = $Z$

-   `pca$rotation` = $\hat{\Gamma}$

-   `pca$center` = mean $\bar{x}$

The second one is (5.35) X = Z $V^\top$ where:

-   `pca$rotation` = $\hat{\Gamma}̂ = V$ = our loadings

Therefore I multiplied 5.22 with the inverse $\hat{\Gamma}̂^\top$ and it becomes: $Z \hat{\Gamma}̂^\top =X-1\bar{x}^\top$

Since I centered the dataset for PCA, the mean is already included, so now just fixing the way the equation shows:

$$
X = Z \hat{\Gamma}̂^\top +1\bar{x}^\top
$$ 

This is done by `xhat`. In the reconstruct function I had to add a limit on each observation for pixel creation as some estimations were out of [0,1].

```{r}
k <- 10

reconstruct <- function(pca){
  Xhat <- pca$x[, 1:k] %*% t(pca$rotation[, 1:k])
  Xhat <- Xhat + matrix(pca$center, nrow(Xhat), ncol(Xhat), byrow=TRUE)
  Xhat[Xhat < 0] <- 0  #not related to PCA, it is for color intensity [0,1] 
  Xhat[Xhat > 1] <- 1  #color intensity must be [0,1] for picture making 
  return(Xhat)
}

R_rec <- reconstruct(pca_R)
G_rec <- reconstruct(pca_G)
B_rec <- reconstruct(pca_B)

```

Now I copied the x format and named it `x_rec` as my new reconstructed x. Then I simply plug now my new values into it for each color and plot. For k=10 we see that it did capture most of the variation in order to recreate the picture.

```{r}
x_rec <- x
x_rec@red   <- R_rec
x_rec@green <- G_rec
x_rec@blue  <- B_rec

plot(x_rec)
```

## 2.3 Selecting the number of components

Although PC1-PC10 capture 90% of total variance of the xray picture, it is not enough variance (in our case pixel variance) to visualize the details. So so of the details that we are missing, like the metal piece of the index finger, sharpness around the fingers etc, we would need to add additional PCs. To find the correct number of components we would need cumulative variance of the number of PCs to be close to 1. This happens around k=50 and following variance graph for a random color, like red, confirms that with k=50 we nearly approach total variance 99%.

```{r}
par(mfrow = c(1, 2))

plot(cum_var_red[1:50],
     xlab = "746 Principal Components",
     ylab = "Cumulative variance",
     main = "Cum. variance  - red")
abline(h = 0.99, col = "red", lty = 2)

k <- 50

reconstruct <- function(pca){
  Xhat <- pca$x[, 1:k] %*% t(pca$rotation[, 1:k])
  Xhat <- Xhat + matrix(pca$center, nrow(Xhat), ncol(Xhat), byrow=TRUE)
  Xhat[Xhat < 0] <- 0  #not related to PCA, it is for color intensity [0,1] 
  Xhat[Xhat > 1] <- 1  #color intensity must be [0,1] for picture making 
  return(Xhat)
}

R_rec <- reconstruct(pca_R)
G_rec <- reconstruct(pca_G)
B_rec <- reconstruct(pca_B)

x_rec <- x
x_rec@red   <- R_rec
x_rec@green <- G_rec
x_rec@blue  <- B_rec

plot(x_rec)

```

## 2.4 Compression factor

To quantify the storage savings, we compare the size of the original RGB image with its PCA representation. The original image consists of three $1012 \cdot 746$ matrices: 

$$
\text{original size} = 3 \cdot 1012 \cdot 746
$$

After PCA compression, each color channel requires $1012k$ scores, $746k$ loadings, and $746$ means. Thus the total number of stored values is 

$$
\text{size}_{\text{PCA}} = 3 \bigl( 1012k + 746k + 746 \bigr).
$$

The compression factor is therefore

$$
\text{compression factor}(k)
= \frac{1012 \cdot 746}{k(1012 + 746) + 746}
$$

For example, with k = 20 principal components we get CF(20) = 21  meaning that the PCA representation requires about 21 times less storage than the full image.

```{r}
CF <- function(k){(1012 * 746) / (k * (1012 + 746) + 746)}

CF(20)
```
