---
title: "Robust regression for college acceptance prediction"
subtitle: "Section 5"
author: "Georgios Papadopoulos"
date: "2025-11-01"
output: github_document
---

*Comparing least-squares and robust regression models for predicting college acceptance rates in R*

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.path = "figures/"
  )
```

# Preparation

After loading and summarizing, I removed the three unwanted variables. Applied log transformation where needed and renamed columns. Additionally I visualize the skewness of before vs after log transforms.

```{r}
library(ISLR)
data("College")
#summary(College)

college_data <- College[, !(names(College) %in% c("Enroll", "Top10perc", "Top25perc"))]

college_data$Accept      <- log1p(college_data$Accept)
college_data$Apps        <- log1p(college_data$Apps)
college_data$F.Undergrad <- log1p(college_data$F.Undergrad)
college_data$P.Undergrad <- log1p(college_data$P.Undergrad)
college_data$Expend      <- log1p(college_data$Expend)

names(college_data)[names(college_data) == "Accept"]      <- "Accept_log"
names(college_data)[names(college_data) == "Apps"]        <- "Apps_log"
names(college_data)[names(college_data) == "F.Undergrad"] <- "F.Undergrad_log"
names(college_data)[names(college_data) == "P.Undergrad"] <- "P.Undergrad_log"
names(college_data)[names(college_data) == "Expend"]      <- "Expend_log"

par(mfrow = c(6, 2), mar = c(1.5, 2, 2, 1), oma = c(0, 0, 0, 0))

hist(College$Accept, main = "Accept (original)", col = "red")
hist(college_data$Accept_log, main = "Accept (log)", col = "green3")

hist(College$Apps, main = "Apps (original)", col = "red")
hist(college_data$Apps_log, main = "Apps (log)", col = "green3")

hist(College$F.Undergrad, main = "F.Undergrad (original)", col = "red")
hist(college_data$F.Undergrad_log, main = "F.Undergrad (log)", col = "green3")

hist(College$P.Undergrad, main = "P.Undergrad (original)", col = "red")
hist(college_data$P.Undergrad_log, main = "P.Undergrad (log)", col = "green3")

hist(College$Expend, main = "Expend (original)", col = "red")
hist(college_data$Expend_log, main = "Expend (log)", col = "green3")
```

```{r}
set.seed(123)

n <- nrow(college_data)
train_index <- sample(1:n, size = round(2/3 * n))

train <- college_data[train_index, ]
test  <- college_data[-train_index, ]
```

# 1. Full least squares regression model

The OLS regression model was fitted on the training data using all available predictors after log-transforming the skewed variables. With $R^2$ of 0.9618 and high F statistic of 904.1 with small p it confirms that the overall model is highly significant.

-   Among the predictors, the most influential predictors are `Apps_log`, `F.Undergrad_log`, and `Outstate`
-   Significant negative predictors are Books`and`Expend_log\`

The residuals are approximately symmetric around zero, therefore we assume no major violations of model assumptions and no bias.

*Regarding each plot:*
-   The residuals vs fitted plot shows residuals primarily between 0 and ±0.5. The distribution is slightly asymmetric, with more colleges showing small positive residuals and fewer with larger negative residuals. The lower part of the plot is sparser and Harvard University stands out with a residual below –1 which I think means that it accepted fewer students than the model captures.
-   The QQ plot shows that most residuals lie close to the diagonal line which means close to normality. On the upper right tail, there are few to no deviations, whereas on the lower left side several points deviate from the line. These outliers I think represent universities that accepted far fewer students than the regression captures. The most pronounced outlier is Harvard University which means that this University's acceptance behavior differs a lot from the regression model.
-   The scale location plot shows a good spread of points that are are evenly scattered without a clear increasing or decreasing pattern. Therefore homoscedasticity holds.
-   The Residuals vs Leverage plot shows most colleges have small residuals, therefore there is a dense cluster. A few high-leverage cases on the right side stand out. Harvard University appears on the lower-left side which means low leverage but a large negative residual. This means that Harvard’s predictor values are not extreme compared to other colleges, yet the model substantially overestimates its number of accepted students.

```{r}
model_ls <- lm(Accept_log ~ ., data = train)

summary(model_ls)

par(mfrow = c(2, 2), mar = c(1.5, 2, 2, 1), oma = c(0, 0, 0, 0))
plot(model_ls)
```

# 2. Full MM-regression model

The MM regression achieved a robust residual standard error of 0.1369 and an $R^2$ of 0.9767, slightly better than the LS model. Most predictors kept their direction and significance. `Apps_log` remains the dominant predictors while others became less significant. Eight observations incl. Harvard university received zero robustness weights which means that they were completely excluded from influencing the fit. This highlights the advantage of MM-regression, which stabilizes coefficient estimates by downweighting such outliers.

```{r}
rownames(train)[c(29,49,61,158,197,261,463,476)]
```

-   The standardized residuals vs robust mahalanobis distances plot shows that most colleges are close to the center, meaning their predictor values and residuals are typical. Outliers like Harvard and Yale were downweighted by the robust method, so they no longer affect the model.
-   The Normal QQ plot for the MM regression shows a much narrower range of residuals compared to the LS model. Outliers such as Harvard and Yale which previously had extreme negative residuals around –4, now appear between –0.5 and –1. This indicates that the robust estimation successfully reduced their influence
-   The response vs fitted plot shows most colleges lying close to the diagonal line therefore the predicted and observed log-acceptance values match closely. A few institutions, such as Harvard and Yale, fall below the line, meaning the model overestimates their acceptances.
-   The squared residuals vs fitted values plot, also known as the scale location plot, shows a horizontal pattern that the variance of residuals remains constant across fitted values. The squared residuals on the y axis from the MM regression are smaller by 1 compared to the OLS regression.

```{r}
library(robustbase)
model_mm <- lmrob(Accept_log ~ ., data = train)
summary(model_mm)

par(mfrow = c(2, 3), mar = c(1.5, 2, 2, 1), oma = c(0, 0, 0, 0))
plot(model_mm)
```

# 3. Residuals and robustness weights

The residuals vs robustness weights plot shows that most observations have weights near 1, meaning they were fully included in the MM regression estimation. The 8 universities with weights of 0 show they were identified as strong outliers. These 8 universities show large residuals but have been effectively downweighted to zero.

```{R}
plot(model_mm$residuals, model_mm$rweights,
     xlab = "residuals",
     ylab = "robustness weights",
     main = "residuals vs robustness weights (MM regression)",
     pch = 20)
abline(h = 1, col = "red", lty = 2)
abline(h = 0, col = "blue", lty = 2)
```

To doublecheck that the identified outliers matched the observations with a robustness weight of zero, I created a residual weight table to confirm the weights equal to zero. The points on the blue line represent these observations.

```{r}
library(tidyverse)
#head(cbind(model_mm$residuals, model_mm$rweights))

cbind(residuals = model_mm$residuals, weights = model_mm$rweights) %>%
  as.data.frame() %>%
  filter(model_mm$rweights == 0)
```

# 4. Limitations of LTS regression

The LTS regression method aims to obtain a robust fit by minimizing the sum of the smallest squared residuals, basically ignoring extreme obs. This makes it less sensitive to outliers than OLS. But when attempting to apply the ltsReg() function to the full multivariate model in the College dataset, the estimation fails with the error for following code:
```{r, eval=FALSE}
model_lts_full <- ltsReg(Accept_log ~ ., data = train)
```

> Error in rdiag(x, id.n = id.n, ...) : The MCD covariance matrix was singular.

This issue arises because ltsReg() relies on the Minimum Covariance Determinant (MCD) estimator to identify a robust subset of the data. When the design matrix includes many predictors or when some of them are strongly correlated, as is the case here (e.g. Apps_log and F.Undergrad_log are highly related), the covariance matrix of certain subsets becomes singular and not invertible. Consequently, the algorithm cannot compute the regression coefficients and the estimation process stops.

```{r}
numeric_vars <- train[, sapply(train, is.numeric)]
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
library(corrplot)
corrplot(cor_matrix, method = "color", tl.col = "black", tl.srt = 45)
```

To make ltsReg() work, the model must be simplified to avoid multicollinearity and reduce dimensionality. I only used the most dominant predictor `Apps_log` that allows LTS regression to run. Simplifying the model ensures that the covariance matrix can be inverted.

```{r}
model_lts_simple <- ltsReg(Accept_log ~ Apps_log , data = train)
```

# 5. Expert-specified acceptance model

Both models yield nearly identical conclusions, but the MM regression provides a cleaner fit by ignoring extreme outliers. The LS model achieved $R^2$ =0.958 with a residual standard error of 0.197. The MM regression slightly improved the fit with $R^2$ = 0.976, residual error = 0.137 and automatically downweighted the outlying colleges (e.g., Harvard, Yale, Duke) to reduce their influence on the results. The significant predictors were consistent across both methods with `Apps_log` and `F.Undergrad_log` showing the strongest positive effects on acceptance rates.

```{r}
model_ls_expert <- lm(Accept_log ~ Private + Apps_log + F.Undergrad_log + Outstate + Room.Board + Grad.Rate, data = train)
summary(model_ls_expert)
```

```{r}
model_mm_expert <- lmrob(Accept_log ~ Private + Apps_log + F.Undergrad_log + Outstate + Room.Board + Grad.Rate, data = train)
summary(model_mm_expert)
```

# 6. Cross validation with standard prediction error

The MM regressions provide slightly more robust fits to the data, but the Expert LS and Expert MM models achieve comparable accuracy with less complexity. This makes the expert models more efficient. All four models perform similarly in terms of predictive accuracy, with only minor differences in their median errors and variability. Both expert models despite using fewer predictors, show very compact error distributions and stable results across repetitions. This demonstrates that reducing the number of predictors does not harm predictive performance.

```{r}
library(cvTools)
set.seed(123)

cv_ls         <- cvFit(lm,   Accept_log ~ .  , 
                       data = train,    y = train$Accept_log, K = 10, R = 50)
cv_mm         <- cvFit(lmrob,Accept_log ~ .  , 
                       data = train,    y = train$Accept_log, K = 10, R = 50)
cv_ls_expert  <- cvFit(lm,   Accept_log ~ Private + Apps_log + F.Undergrad_log + Outstate + Room.Board + Grad.Rate, 
                       data = train, y = train$Accept_log, K = 10, R = 50)
cv_mm_expert  <- cvFit(lmrob,Accept_log ~ Private + Apps_log + F.Undergrad_log + Outstate + Room.Board + Grad.Rate, 
                       data = train, y = train$Accept_log, K = 10, R = 50)

cv_errors <- data.frame(
  LS_Full    = cv_ls$reps,
  MM_Full    = cv_mm$reps,
  LS_Expert  = cv_ls_expert$reps,
  MM_Expert  = cv_mm_expert$reps
)

boxplot(cv_errors,
        col   = c("lightyellow", "khaki1", "lightyellow", "khaki1"),
        names = c("Full LS", "Full MM", "Expert LS", "Expert MM"),
        ylab  = "Prediction error",
        main  = "10 fold cross validation")

```

# 7. Cross validation with robust trimmed prediction error

I repeated the cross validation with root trimmed mean squared prediction error (RTMSPE) as the cost function. RTMSPE trims a small proportion of extreme prediction errors which makes it less sensitive to outliers. This allows to evaluate the typical predictive performance of each model rather than being influenced by a few extreme cases. So we can compare all models based on a more robust error measure that downweights the effect of outliers in the cross validation results.

This model comparison is more useful because it is more informative and reliable because RTMSPE provides a fairer assessment of model performance when outliers are present such as `Harvard` and `Yale`. From the boxplots both LS models show higher prediction errors and the MM regression models achieve lower and more consistent ones. The two MM models perform similarly but the expert MM model is more compact and narrower which means that its errors are more stable.

```{r}
set.seed(123)

cv_ls_rtmspe         <- cvFit(lm,   Accept_log ~ .  , 
                       data = train,    y = train$Accept_log, K = 10, R = 50, cost = rtmspe)
cv_mm_rtmspe         <- cvFit(lmrob,Accept_log ~ .  , 
                       data = train,    y = train$Accept_log, K = 10, R = 50, cost = rtmspe)
cv_ls_expert_rtmspe  <- cvFit(lm,   Accept_log ~ Private + Apps_log + F.Undergrad_log + Outstate + Room.Board + Grad.Rate, 
                       data = train, y = train$Accept_log, K = 10, R = 50, cost = rtmspe)
cv_mm_expert_rtmspe  <- cvFit(lmrob,Accept_log ~ Private + Apps_log + F.Undergrad_log + Outstate + Room.Board + Grad.Rate, 
                       data = train, y = train$Accept_log, K = 10, R = 50, cost = rtmspe)

cv_errors_rtmspe <- data.frame(
  LS_Full    = cv_ls_rtmspe$reps,
  MM_Full    = cv_mm_rtmspe$reps,
  LS_Expert  = cv_ls_expert_rtmspe$reps,
  MM_Expert  = cv_mm_expert_rtmspe$reps
)

boxplot(cv_errors_rtmspe,
        col   = c("wheat", "tan", "wheat", "tan"),
        names = c("Full LS", "Full MM", "Expert LS", "Expert MM"),
        ylab  = "Prediction error",
        main  = "10 fold cross validation with RTMSPE")

```

# 8. Test set prediction and graphical evaluation

The plots for all models show a linear relationship between observed and predicted values, so we see that the models capture the main trend well. Both MM regression models show tighter clustering of points around the diagonal and fewer extreme deviations which confirms their robustness to outliers. In contrast, the LS models show slightly greater dispersion due to outliers.

The MM regression models yield more stable and reliable predictions. The expert MM model even with fewer predictors performs almost as well as the full MM model and better than both LS versions. This confirms that the few predictors are enough for accurate and robust prediction.

```{r}
pred_ls        <- predict(model_ls, newdata = test)
pred_mm        <- predict(model_mm, newdata = test)
pred_ls_expert <- predict(model_ls_expert, newdata = test)
pred_mm_expert <- predict(model_mm_expert, newdata = test)

pred_df <- data.frame(
  Observed = test$Accept_log,
  LS_Full = pred_ls,
  MM_Full = pred_mm,
  LS_Expert = pred_ls_expert,
  MM_Expert = pred_mm_expert
)

par(mfrow = c(2, 2))
plot(pred_df$Observed, pred_df$LS_Full, 
     main = "Full LS Model", xlab = "Observed", ylab = "Predicted", pch = 19, col = "green3")
abline(0, 1, lwd = 1.5, col = "red")

plot(pred_df$Observed, pred_df$MM_Full, 
     main = "Full MM Model", xlab = "Observed", ylab = "Predicted", pch = 19, col = "green3")
abline(0, 1, lwd = 1.5, col = "red")

plot(pred_df$Observed, pred_df$LS_Expert, 
     main = "Expert LS Model", xlab = "Observed", ylab = "Predicted", pch = 19, col = "green3")
abline(0, 1, lwd = 1.5, col = "red")

plot(pred_df$Observed, pred_df$MM_Expert, 
     main = "Expert MM Model", xlab = "Observed", ylab = "Predicted", pch = 19, col = "green3")
abline(0, 1, lwd = 1.5, col = "red")

```
