---
title: "UE Multivariate Statistics"
subtitle: "Section 9"
author: "Georgios Papadopoulos | "
date: "2025-12-01"
output: github_document
---

*Assessing relationships between relative weight and glucose-insulin features across diagnostic groups*

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.path = "figures/"
  )
```

# 1. Multiple correlation by diagnostic group

The Diabetes dataset consists of p=6 variables and n=145 obs, out of which the ones related with glucose plasma are assigned onto `gluc.vars`. There are three levels of groups, namely Normal, Chemical_Diabetic and Overt_Diabetic.

```{r}
library(heplots)

data("Diabetes")
#?Diabetes

str(Diabetes)
summary(Diabetes)

gluc.vars <- c("glufast", "glutest", "instest", "sspg")
groups <- levels(Diabetes$group)
```

To calculate r for each group I wrote a function that calculates the correlation matrix for each group from page 100/126.

$$
r^{2}_{x,y} = R^{\top}_{yx} \, R^{-1}_{yy} \, R_{yx}
$$
Where
x = relwt and 
y = (glufast, glutest, instest, sspg)

```{r}
compute_multcorr <- function(g) {
  group_data <- subset(Diabetes, group == g)
  
  x <- group_data$relwt
  y <- group_data[, gluc.vars]
  
  Ryy <- cor(y)
  Ryx <- cor(y, x)
  
  r2_value <- as.numeric(t(Ryx) %*% solve(Ryy) %*% Ryx)
  r_value  <- sqrt(r2_value)
  
  data.frame(group = g, r2 = r2_value, r = r_value)
}

ex1_results <- do.call(rbind, lapply(groups, compute_multcorr))
```

Across the diagnostic groups, the multiple correlation increases from 0.55 in the normal group to 0.56 in the chemical diabetic group and reaches 0.72 in the overt diabetic group. This means that body weight is much more closely related to glucose levels in people with more advanced diabetes.

```{r}
ex1_results
```

# 2. Interpretation of linear predictor coefficients

The coefficients of the linear predictor $a_{0}+a^{\top}y$ can be interpreted in the same way as regression coefficients. Each coefficient describes how the predicted relative weight `relwt` changes when one glucose plasma variable increases, while the other variables are kept fixed. A positive coefficient means that higher values of this variable are associated with higher predicted relative weight `relwt` and a negative coefficient means the opposite. The larger the absolute value of a coefficient, the more important that variable is in the linear prediction of `relwt`. we can see which glucose variables are most influential in each group and how their influence changes with group severity by comparing these coefficients between groups. 

Because I was not sure from the exercise instruction whether I needed to compute the coefficients, I did it depending on each of the three groups as we saw for sample data in the lecture with $a=R^{-1}_{yy}\,R_{yx}$ where we see for example that for every group the `sspg` variable has the most influence on each group with a positive coeff.

```{r}
compute_coeffs <- function(g) {
  group_data <- subset(Diabetes, group == g)

  x <- group_data$relwt
  y <- group_data[, gluc.vars]

  Ryy <- cor(y)
  Ryx <- cor(y, x)

  a  <- solve(Ryy) %*% Ryx
  rownames(a) <- gluc.vars

  data.frame(group = g, variable = gluc.vars, coefficient = as.numeric(a))
}

coeffs_ex2 <- do.call(rbind, lapply(groups, compute_coeffs))

coeffs_ex2

```


# 3. Significance of multiple correlations

To test whether the multiple correlation is zero, we use the F-test

$$
F = \frac{(n - 1 - p)\, r^{2}_{x,y}}{p(1 - r^{2}_{x,y})}
$$


with p=4 predictors and n=145 obs.

Codewise, I do the same function as in task 1 but now I also calculate Fstat and the right tailed probability.

```{r}
compute_test <- function(g) {
  group_data <- subset(Diabetes, group == g)

  x <- group_data$relwt
  y <- group_data[, gluc.vars]

  Ryy <- cor(y)
  Ryx <- cor(y, x)

  r2 <- as.numeric(t(Ryx) %*% solve(Ryy) %*% Ryx)
  r  <- sqrt(r2)

  n  <- nrow(y)
  p  <- ncol(y)

  Fstat <- (n - 1 - p) * r2 / (p * (1 - r2))
  pval  <- 1 - pf(Fstat, p, n - 1 - p)

  data.frame(
    group = g,
    r = r,
    r2 = r2,
    F = Fstat,
    p.value = pval
  )
}

ex3_results <- do.call(rbind, lapply(groups, compute_test))
```

Although a relatively large $r^2$ already suggests that the multiple correlation is likely to be significant, statistical significance cannot be determined from  $r^2$  alone, because it also depends on the sample size and the number of predictors. The Ftest confirms that all three  $r^2$ values are significantly different from zero, as all p-values fall below the 0.05 threshold. Therefore, we reject the null hypothesis that the glucose plasma variables are jointly uncorrelated with weight in every group.

```{r}
ex3_results
```

# 4. Pearson CCA grid comparison

Same results with CCAgrid. This happens because we only have one variable in X = `relwt`. In that case, the canonical correlation is the same thing as the multiple correlation. So CCAgrid gives the same numbers as before.


```{r}
library(ccaPP)
#?CCAgrid

# Subset group 1
g1 <- subset(Diabetes, group == "Normal")

X <- as.matrix(g1$relwt)
Y <- as.matrix(g1[, gluc.vars])

res1 <- CCAgrid(X, Y, k = 1, method = "pearson", standardize = TRUE)

# Subset group 2
g2 <- subset(Diabetes, group == "Chemical_Diabetic")

X <- as.matrix(g2$relwt)
Y <- as.matrix(g2[, gluc.vars])

res2 <- CCAgrid(X, Y, k = 1, method = "pearson", standardize = TRUE)

# Subset group 3
g3 <- subset(Diabetes, group == "Overt_Diabetic")

X <- as.matrix(g3$relwt)
Y <- as.matrix(g3[, gluc.vars])

res3 <- CCAgrid(X, Y, k = 1, method = "pearson", standardize = TRUE)

ex4_results <- c(normal = res1$cor, chemical = res2$cor, overt = res3$cor)
ex4_results
```

# 5. Spearman CCA grid comparison

method = "spearman" produces slightly higher correlations than Pearson, but the overall pattern remains the same. The methodological difference is that Spearman correlation is rank based and captures also non linear relationships, while Pearson correlation measures linear relationships using the raw data. Spearman is therefore less sensitive to outliers, which I believe explains the small increase in the correlation values.

```{r}

# Subset group 1
g1 <- subset(Diabetes, group == "Normal")

X <- as.matrix(g1$relwt)
Y <- as.matrix(g1[, gluc.vars])

res1 <- CCAgrid(X, Y, k = 1, method = "spearman", standardize = TRUE)

# Subset group 2
g2 <- subset(Diabetes, group == "Chemical_Diabetic")

X <- as.matrix(g2$relwt)
Y <- as.matrix(g2[, gluc.vars])

res2 <- CCAgrid(X, Y, k = 1, method = "spearman", standardize = TRUE)

# Subset group 3
g3 <- subset(Diabetes, group == "Overt_Diabetic")

X <- as.matrix(g3$relwt)
Y <- as.matrix(g3[, gluc.vars])

res3 <- CCAgrid(X, Y, k = 1, method = "spearman", standardize = TRUE)

ex5_results <- c(normal = res1$cor, chemical = res2$cor, overt = res3$cor)
ex5_results

```

# 6. Effect of transformations on rank-based results

Spearman correlation is based on ranks, not on the original numerical values. Therefore, monotone transformations such as the logarithm should not change the rank order of the observations and should not change the Spearman correlations.

In this setting, `method = "spearman"` applies the CCA procedure to rank-based information. Since the log transformation preserves the ordering of the variables, the results should remain the same or nearly identical, apart from possible numerical differences.

Therefore, transforming the variables before applying Spearman-based CCA is generally unnecessary. Transformations are more relevant for Pearson-based methods, where the actual values, scale, skewness, and outliers directly affect the estimated correlations.

```{r}
Diabetes_log <- Diabetes
Diabetes_log$relwt   <- log(Diabetes$relwt + 1)
Diabetes_log$glufast <- log(Diabetes$glufast + 1)
Diabetes_log$glutest <- log(Diabetes$glutest + 1)
Diabetes_log$instest <- log(Diabetes$instest + 1)
Diabetes_log$sspg    <- log(Diabetes$sspg + 1)

g1 <- subset(Diabetes_log, group == "Normal")
X1 <- as.matrix(g1$relwt)
Y1 <- as.matrix(g1[, c("glufast","glutest","instest","sspg")])

res1 <- CCAgrid(X1, Y1, k = 1, method = "spearman", standardize = TRUE)
res1$cor

g2 <- subset(Diabetes_log, group == "Chemical_Diabetic")
X2 <- as.matrix(g2$relwt)
Y2 <- as.matrix(g2[, c("glufast","glutest","instest","sspg")])

res2 <- CCAgrid(X2, Y2, k = 1, method = "spearman", standardize = TRUE)
res2$cor

g3 <- subset(Diabetes_log, group == "Overt_Diabetic")
X3 <- as.matrix(g3$relwt)
Y3 <- as.matrix(g3[, c("glufast","glutest","instest","sspg")])

res3 <- CCAgrid(X3, Y3, k = 1, method = "spearman", standardize = TRUE)
res3$cor

```

# 7. Permutation test for uncorrelatedness

permTest does a permutation test for uncorrelatedness by repeatedly shuffling the rows of the first input X. Each shuffle breaks any real relationship between X and Y. After each permutation, the function creates a new canonical correlation. These correlations represent what the correlation would look like if X and Y were unrelated. Then we compare the real correlation to these fake ones basically. If the real correlation is bigger than all of the permuted ones, it means the relationship is real and not just random. If X and Y were truly unrelated, shuffling X would not change anything. A permutation test does not rely on a distributional assumption.

# 8. Pearson permutation test results

The permutation test shows that all three canonical correlations are statistically significant.

 - For the Normal group, the p-value is essentially 0, indicating very strong evidence of association.

- For the Chemical Diabetic group, the p-value is 0.03, which is still below 0.05, so the association is also significant, although weaker.

- For the Overt Diabetic group, the p-value is again effectively 0, showing a very strong relationship.

These results match the earlier F tests from Exercise 3: in every group, the glucose plasma variables are significantly related to `relwt`.

```{r}
#?permTest
# subset group 1
g1 <- subset(Diabetes, group == "Normal")

X1 <- as.matrix(g1$relwt)             # first set
Y1 <- as.matrix(g1[, gluc.vars])      # second set

set.seed(123)
perm_normal <- permTest(X1, Y1, R = 1000, method="pearson")
perm_normal

# subset group 2
g2 <- subset(Diabetes, group == "Chemical_Diabetic")
X2 <- as.matrix(g2$relwt)
Y2 <- as.matrix(g2[, gluc.vars])

set.seed(123)
perm_chemical <- permTest(X2, Y2, R = 1000, method="pearson")
perm_chemical

# subset group 3
g3 <- subset(Diabetes, group == "Overt_Diabetic")
X3 <- as.matrix(g3$relwt)
Y3 <- as.matrix(g3[, gluc.vars])

set.seed(123)
perm_overt <- permTest(X3, Y3, R = 1000, method="pearson")
perm_overt
```

# 9. Spearman permutation test results

The Spearman permutation results are very similar to the Pearson permutation results. This is because permTest() performs canonical correlation analysis CCA, not simple correlation. CCA finds the strongest possible linear association between the variable sets, and this optimization step makes the result insensitive to whether Pearson or Spearman correlations were used. Therefore in this case both methods give identical permutation results, although methodologically are different.

```{r}
# subset group 1
g1 <- subset(Diabetes, group == "Normal")

X1 <- as.matrix(g1$relwt)             # first set
Y1 <- as.matrix(g1[, gluc.vars])      # second set

set.seed(123)
perm_normal <- permTest(X1, Y1, R = 1000, method="spearman")
perm_normal

# subset group 2
g2 <- subset(Diabetes, group == "Chemical_Diabetic")
X2 <- as.matrix(g2$relwt)
Y2 <- as.matrix(g2[, gluc.vars])

set.seed(123)
perm_chemical <- permTest(X2, Y2, R = 1000, method="spearman")
perm_chemical

# subset group 3
g3 <- subset(Diabetes, group == "Overt_Diabetic")
X3 <- as.matrix(g3$relwt)
Y3 <- as.matrix(g3[, gluc.vars])

set.seed(123)
perm_overt <- permTest(X3, Y3, R = 1000, method="spearman")
perm_overt
```


