---
title: "Rotated PCA and factor analysis of handwriting features"
subtitle: "Section 8"
author: "Georgios Papadopoulos | "
date: "2025-11-22"
output: github_document
---

*Comparing rotated principal components and latent factor structures in handwriting measurements*

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.path = "figures/"
  )
```

# 1. Rotated principal components

The darwinM dataset contains 19 handwriting measurements collected from two groups/classes of participants: healthy people `H` and patients diagnosed with Alzheimer’s `P`. Each participant completed 25 standardized writing tasks and our data has the median per measurement/variable. The dataset includes variables describing timing, speed, acceleration, pressure, jerk etc.

With `summary(PCA)` I can explore that cumulative proportion of PC1 and PC2 is 0.61, meaning that the first two components account for 61% of total variance in the data. So a large part of the multivariate dataset can be represented in a two-dimensional PCA plot, making PC1 and PC2 suitable for visualizing.

```{r}
load("darwinM.RData")

X <- darwinM[, -19]

pca <- princomp(X, cor = TRUE, center = TRUE)

summary(pca)
```

## 1.1 Selecting the number of components

I selected k=3 principal components so the cumulative explained variance is more than 70% because rotation should only be applied to the meaningful part of the PCA solution. 60% variance with the first 2 PCs leaves a lot of information out and leads to incomplete rotated result.

## 1.2 Varimax rotation of PCA loadings

I extracted the unrotated loadings of the first k components and applied varimax rotation to make the structure easier to interpret. 
Varimax rotation makes the components easier to interpret by pushing loadings either very high or near zero. That's why each variable ends up clearly associated with one specific component instead of being spread across several. 

`varimax_res`contains `loadings` and `rotmat` rotation matrix

Then I created the rotated scores by multiplying the PCA scores with the rotation matrix.

```{r}
k <- 3

loadings_unrot <- unclass(pca$loadings)[, 1:k]

varimax_res <- varimax(loadings_unrot)

loadings_rot <- varimax_res$loadings
```

## 1.3 Rotation of PCA scores

I applied the same orthogonal rotation matrix used for the loadings to rotate the PC scores.

Then I multiplied the PCA scores with the rotation matrix.

```{r}
rotation_matrix <- varimax_res$rotmat
scores_rot <- pca$scores[, 1:k] %*% rotation_matrix
```

## 1.4 Biplot comparison of rotated and unrotated PCA

In the unrotated PCA biplot, PC1 captures the largest proportion of variance and PC2 captures the second largest, but the loading structure is difficult to interpret because many variables have moderate contributions to both dimensions. For example pressure_mean loads heavily on PC1 but only weakly on PC2. Another example is max_y_extension because it contributes strongly to PC2 and nothing to PC1.

Before rotation:

- PC1 explains 44% variance

- PC2 explains 16.7% variance

These percentages come directly from the eigenvalues of PCA. After rotation the rotation redistributes the variance among the k PCs.

After rotating here for k = 3 the rotation tries to even out and simplify the loadings. The percentages on the rotated axes refer to how much of the total variance of k components is explained by RC1 and RC2

- Rotated PC1 = 61.9%

- Rotated PC2 = 23.5%

- Rotated PC3 = 100% - (61.9% + 23.5%) = 14.6% since k = 3

The loadings become clearer because variables cluster more distinctly along the rotated axes. Now pressure_mean shows a more balanced influence on both RC1 and RC2. Similarly, max_y_extension still loads strongly on RC1 but I can now also see a loading on the second rotated component. This reveals that within the selected k = 3 subspace, max_y_extension contributes primarily to one latent dimension but also plays a secondary role in another.

I used `repel = TRUE` so that variable names in the biplot don't overlap.

```{r}
library(factoextra)

fviz_pca_biplot(
  pca,
  repel = TRUE,
  title = "Unrotated PCA biplot"
)

pca_rotated <- list(
  x = scores_rot,     
  rotation = loadings_rot,
  center = pca$center,
  scale = pca$scale,
  sdev = pca$sdev[1:k]
)

class(pca_rotated) <- "prcomp"     # trick fviz_pca_biplot into accepting it

fviz_pca_biplot(
  pca_rotated,
  repel = TRUE,
  title = "Rotated PCA biplot"
)
```

# 2. Factor analysis of handwriting features

I selected k=3 because before I found out that 3 PCs explain 70% cumulative variance of the data which is sufficient.

## 2.1 Factor model estimation

The factor analysis with 3 factors explains 63.1% of the total variance. Variables with low values like air_time, gmrt_on_paper, mean_acc_in_air  and in-air jerk total_time have low uniqueness. That means the factor analysis explain these variables well. Variables with high uniqueness are less related to the 3 factors.

The chi-square test indicates that 3 factors do not fully reproduce the correlation matrix.

```{r}
k <- 3

fa <- factanal(scale(X), factors = k, scores = "regression")

fa
```

## 2.2 Maximum number of factors

The largest number of factors I could technically use is the number of variables minus one. Since we have 18 variables because I removed `class`, the maximum possible number of factors is k = 17.

However large k don't make sense in practice. Factor analysis is only useful with a small number of factors, usually around k=5.

## 2.3 PCA and factor analysis comparison

In PCA, the loadings show how much each variable contributes to the principal components, which represent directions of maximum variation in the data. 
In factor analysis, the loadings show how strongly each variable relates to hidden factors that explain the common patterns in the dataset.
So basically PCA looks at total variance, while factor analysis looks only at shared variance.

The scores also differ. PCA scores come from projecting each observation onto the principal components. Factor scores are estimates of how much each observation expresses each latent factor. They depend on the statistical model and the method used to compute them.

Uniquenesses are special to factor analysis. They show how much of a variable’s variance the factors cannot explain. I already described them in section 2.1 but basically:

- A low uniqueness means the factors explain the variable well closer to 0. 

- A high uniqueness means the variable mostly contains its own specific variation or noise closer to 1. 

## 2.4 Factor biplot and interpretation

Similar to PCA, I assign the factor scores and factor loadings for the first two factors for the biplot. I put them in a list structure so visualization works. Compared to PCA the first two factors from factor analysis are easier to interpret.

- Factor 1 has strong loadings on air_time, paper_time and total_time. It represents the time a person writes. 

- Factor 2 has high loadings on gmrt_on_paper, mean_speed_on_paper, and the extension variables. It shows how fast a person writes. PCA combined these patterns more loosely, while factor analysis separates them into two clear underlying dimensions.

```{r}
FA_scores   <- fa$scores[, 1:2]
FA_loadings <- fa$loadings[, 1:2]

fa_biplot_obj <- list(
  sdev     = apply(FA_scores, 2, sd),
  rotation = FA_loadings,
  center   = FALSE,
  scale    = FALSE,
  x        = FA_scores
)
class(fa_biplot_obj) <- "prcomp" #trick fviz_pca_biplot into accepting it

fviz_pca_biplot(fa_biplot_obj, repel = TRUE, title = "Factor analysis biplot")
```


