---
title: "Cluster validation for wine profiling"
subtitle: "Section 3"
author: "Georgios Papadopoulos"
date: "2025-10-17"
output: github_document
---

*Evaluating K-Means, Hierarchical, Fuzzy, and Model-Based Clustering with Internal Validity Measures in R*

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/"
  )
```

# Preparation

```{r}
library(doBy)
data(wine)
wine_numbers <- wine[, sapply(wine, is.numeric)]
wine_scaled <- scale(wine_numbers)
```

# 1. Calinski-Harabasz and Hartigan indices for K-means

To calculate Calinksi Harabasz index I used `fpc` package <https://search.r-project.org/CRAN/refmans/fpc/html/calinhara.html> where I double checked that we use the same formula.

We calculate the between-cluster variance $B_k$ divided by the within-cluster variance $W_k$ then take the natural logarithm:

$$
H_k=ln\frac{W_k}{B_k}
$$

The optimal cluster k=3 since for the Calinski Harabasz index we are interested in the biggest value which we see both visually and on the data for k=3. For the Hartigan index we are interested in the largest change of Hartigan, namely $ΔH_k=H_k−H_{k−1}$ which happens for k=3. Basically the biggest increase is when we approach when from cluster 2 we add **cluster 3**. From our previous exercise, we know that we had three types of wines.

```{r}
library(fpc)
library(cluster)
library(dplyr)

data(wine)
wine_numbers <- wine[, sapply(wine, is.numeric)]
wine_scaled <- scale(wine_numbers)

set.seed(123)
k <- 2:10
CH_index <- numeric(length(k))
Hartigan_index <- numeric(length(k))


for (i in seq_along(k)) {
  res <- kmeans(wine_scaled, centers = k[i], nstart = 25)
  
  CH_index[i] <- calinhara(wine_scaled, res$cluster)
  
  Hartigan_index[i] <- log(res$betweenss / res$tot.withinss)
}


results_task1 <- data.frame(
  K = k,
  Calinski_Harabasz = CH_index,
  Hartigan = Hartigan_index
)

print(results_task1)

par(mfrow = c(1, 2))

plot(k, CH_index, type = "b", pch = 20,
     xlab = "Number of clusters",
     ylab = "index value",
     main = "Calinski Harabasz Index")

plot(k, Hartigan_index, type = "b", pch = 20,
     xlab = "Number of clusters",
     ylab = "index value",
     main = "Hartigan Index")


```

# 2. Silhouette analysis for k-means

Αverage silhouette width measures how well each observation fits within its cluster compared to other clusters. On the numerator there is the difference of smallest average distance to points in the nearest cluster with the average distance to all points in its own cluster c.

$$
s_i = \frac{d_{i,c} - d_{i,ck}}{max(d_{i,ck},d_{i,c})}
$$ 
Then we take the mean of column 3 to obtain the average silhouette width which summarizes the overall clustering quality for each cluster. I tested this before with `head(sil)`.

Although the values are relatively low the highest average is for k = 3 which means that three clusters provide the best clustering structure. The values are low because we remember from the last exercise that there was a significant overlap throughout two of three clusters `plot(wine, col = wine$Cult, pch = 16)`

```{r}
# Initialize
set.seed(123)
k <- 2:10
silhouette_values <- numeric(length(k))


for (i in seq_along(k)) {
  km <- kmeans(wine_scaled, centers = k[i], nstart = 25)
  sil <- silhouette(km$cluster, dist(wine_scaled))
  silhouette_values[i] <- mean(sil[, 3])  # column 3 = silhouette width
}


results_sil <- data.frame(k, silhouette_values)
print(results_sil)

plot(k, silhouette_values, type = "b", pch = 20,
     xlab = "Number of clusters",
     ylab = "Average silhouette width",
     main = "Average silhouette width for K-means clustering")

```

# 3. Silhouette analysis for hierarchical clustering

Now we will use again the average silhouette width as a validity measure, but change the clustering method. Instead of kmeans we use hierarchical clustering.

Again the highest average silhouette value is at K = 3, meaning that the clustering quality is then the best. The results are consistent with k-means clustering. Between the two algorithms k-means achieved slightly higher silhouette values which means slightly better cluster separation.

There is a small rise after k=6 which is likely because the existing groups are split into smaller parts rather than forming new meaningful clusters.

```{r}
dist_matrix <- dist(wine_scaled)

hc <- hclust(dist_matrix, method = "ward.D2")

# initialize
k <- 2:10
silhouette_values_hc <- numeric(length(k))

for (i in seq_along(k)) {
  clusters <- cutree(hc, k[i])
  sil <- silhouette(clusters, dist_matrix)
  silhouette_values_hc[i] <- mean(sil[, 3])
}


data.frame(K = k, Avg_Silhouette = silhouette_values_hc)


plot(k, silhouette_values_hc, type = "b", pch = 20,
     xlab = "Number of clusters",
     ylab = "Average silhouette width",
     main = "Average silhouette width for hierarchical clustering")


```

# 4. Gap atatistic for k-means

The Gap statistic compares how compact clusters are with what would happen in random data that has no real clusters. A big Gap value means clusters are much more compact than random noise. The best number of clusters is the smallest k where the Gap value is almost as high as the next one.

The Gap statistic rises until k = 3 with Gap = 0.63 and then stays almost the same. Since the standard errors are very small the difference after k = 3 is not meaningful.
According to the standard error rule the best number of clusters is k = 3.

The nstart parameter in k-means specifies how many times the k-means algorithm should run with random initial centroids. With nstart = 25+ we ensure that k-means explores multiple starting points and chooses the best clustering result.

```{r}
set.seed(123)
gap <- clusGap(wine_scaled,
                  FUN = kmeans,
                  K.max = 10,
                  nstart = 25,
                  B = 50) #number of random bootstrap samples

gap

plot(gap, main = "Gap statistic with k means clustering")

opt_k <- maxSE(gap$Tab[, 3], gap$Tab[, 4])
opt_k
```

# 5. Gap statistic for fuzzy clustering

The difference with the previous exercise is that now we implement fuzzy cmeans clustering. m is a parameter than controls how fuzzy the clustering is. For m=1 clustering is exactly like k-means, but as m increases the points can belong to multiple clusters.

For our exercise the Gap statistic increases up to K = 3, after which it stabilizes. This confirms the previous findings from k-means and hierarchical clustering.


```{r}
library(fclust)
library(e1071)
set.seed(123)

gap_fuzzy <- clusGap(wine_scaled,
                     FUNcluster = cmeans,
                     m = 1.5,         
                     K.max = 10,
                     B = 50)

plot(gap_fuzzy, main = "Gap statistic for fuzzy clustering")

gap_fuzzy

optimal_k_fuzzy <- maxSE(gap_fuzzy$Tab[,3], gap_fuzzy$Tab[,4])
optimal_k_fuzzy
```

# 6. Efficient gap statistic for model based clustering

To use the Gap statistic for model-based-clustering, at first I used the Bayesian Information Criterion (BIC) to find the best model which is VVE for variable volume, variable shape and equal orientation. Basically different sizes and shapes but all elipsoids tilted the same way. In this way I will calculate the Gap statistics only for VVE.

```{r}
library(mclust)

bic <- mclustBIC(wine_scaled)
#summary(bic)                
best_model <- Mclust(wine_scaled)  # Fits the best automatically
best_model$modelName
```
Mclust() tries different numbers of clusters G=1,2,...,10 and chooses the one with the best BIC. This is to briefly check if we will get the optimal number of clusters with VVE model. Once again k=3 so now I will calculate Gap.

```{r}
Mclust(wine_scaled, modelNames = "VVE")$G
```

Now I run the Gap statistic only on VVE. For each number of clusters it fits a Gaussian mixture model using the `Mclust()` function.

I used a wrapper around Mclust() because of vector and list issues. If a model cannot be fitted, now all data points are temporarily assigned to one cluster so the process can continue.


```{r}
set.seed(123)

mclustGap <- function(x, k) {
  model <- try(Mclust(x, G = k, modelNames = "VVE", verbose = FALSE), silent = TRUE)
  if (inherits(model, "try-error") || is.null(model$classification)) {
    cl <- rep(1, nrow(x))  # fallback: assign all to one cluster
  } else {
    cl <- model$classification
  }
  list(cluster = cl)
}


gap_mclust <- clusGap(wine_scaled,
                      FUNcluster = mclustGap,
                      K.max = 10,
                      B = 50)


plot(gap_mclust, main = "Gap statistic for VVE Model")


optimal_k <- maxSE(gap_mclust$Tab[,3], gap_mclust$Tab[,4])
optimal_k

```