---
title: "Exploratory Multivariate Analysis of the Iris Dataset"
subtitle: "Exercise 1"
author: "Georgios Papadopoulos"
date: "2025-10-05"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This report presents a compact multivariate analysis of the classic `iris` dataset. The goal is to explore relationships between variables, summarize key statistical properties, 
and verify fundamental results from multivariate theory using empirical data.

# Dataset Overview

We begin by exploring a well-known dataset from the base R datasets package. The most famous R datasets are iris, mtcars, AirPassengers and USArrests. The `iris` dataset contains 150 obs, consists of 4 numeric continuous variables and 1 categorical variable.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(datasets)
#?iris
str(iris)
```

-   The 4 numeric variables are the width and length of specific flower parts in centimeters named sepal and petal. See photo <https://www.sciencefacts.net/sepals.html>
-   The categorical variable is the species of iris flower. The three iris species in the dataset are:

```{r, echo=FALSE}
#iris$Species
unique(iris$Species)
```

# Exploratory Data Analysis

### Distribution of variables with histograms

The histograms show the frequency distributions of all four numeric variables in the iris dataset.

-   Sepal Length and Sepal Width are approximately normally distributed, with most flowers having sepal lengths between 5–6 cm and widths around 3 cm.
-   Petal Length and Petal Width show multiple distributions which could be because of the different species cluster around the distinct size range till 2 cm for petal length.

```{r, echo=FALSE, fig.width=10, fig.height=3}
par(mfrow = c(1, 4))

hist(iris$Sepal.Length, main = "Sepal Length", xlab = "cm")
hist(iris$Sepal.Width, main = "Sepal Width", xlab = "cm")
hist(iris$Petal.Length, main = "Petal Length", xlab = "cm")
hist(iris$Petal.Width, main = "Petal Width", xlab = "cm")
```

### Scatterplot of pairwise relationships

The scatterplot matrix displays all pairwise relationships colored by species. Black for setosa, red forversicolor and green for virginica.

-   Petal length and petal width show a strong positive correlation with clusters that are good separated for each species.
-   Sepal measurements are less strongly correlated and show more overlap between species.
-   Setosa appears clearly distinct from versicolor and virginica, because the other two partially overlap.

```{r, echo=FALSE, fig.width=5, fig.height=5}
par(mar = c(5, 0, 4, 8), xpd = TRUE)  # add right margin & allow drawing outside

plot(iris, col = iris$Species, pch = 16, main = "Scatterplot Matrix of Iris Dataset")

legend(legend = levels(iris$Species), col = 1:3, pch = 1, bty = "n", "bottomright", inset = 0.02)
```

### Boxplot group comparisons

The boxplots show the interquartile range and the line inside the box marks the median value. So we can compare all four numeric variables across the three iris species.

-   Setosa has the smallest values for most measurements.
-   Virginica has the largest while versicolor is in between.
-   The petal length and petal width plots show very clear species separation, with little to no overlap.
-   The sepal part of the flowers overlap more, especially between versicolor and virginica.

```{r, echo=FALSE, fig.width=10, fig.height=3}
par(mfrow = c(1, 4))

boxplot(Sepal.Length ~ Species, data = iris,  main = "Sepal length")
boxplot(Sepal.Width ~ Species, data = iris, main = "Sepal width")
boxplot(Petal.Length ~ Species, data = iris,  main = "Petal length")
boxplot(Petal.Width ~ Species, data = iris,main = "Petal width")
```

# Descriptive Statistics

Summary statistics provide a quantitative view of the variable distributions:
Sepal length and width show smaller ranges between 4-8cm compared to petal length and width which seem from 0.1 till 2.5 cm. This might indicate diversity in petal sizes among species.

```{r, echo=FALSE}
summary(iris)
```
Following table summerizes when we group by species. Length measurement for both petal and sepal increase from setosa to virginica species, showing that petals are the strongest indicators for species classification. However, sepal width differentiates less across groups.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

iris %>%
  group_by(Species) %>%
  summarise(across(where(is.numeric),
                   list(mean = mean, 
                        sd = sd),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(-Species,
               names_to = c("Variable", "Statistic"),
               names_sep = "_",
               values_to = "Value") %>%
  pivot_wider(names_from = Species, values_from = Value)

```


# Covariance and Correlation structure

## Covariance matrix

#### Variances cov(X, X))

The diagonal elements of the covariance matrix represent the variances of each variable, indicating how much each measurement fluctuates around its mean.

-   Petal length shows the largest variance of 3.12, which means that it differs a lot among all iris samples
-   Petal width and Sepal.Length show moderate variation
-   Sepal width shows the smallest variance of 0.19 therefore sepal width is the most consistent characteristic.

#### Covariances cov(X,Y)

Covariances between same flower part: - Sepal length and width show negative covariance of -0.04. It means that longer sepals are very slightly associated with negative sepal width (narrow). - Petal length and width show positive covariance of 1.29. It means that longer petals are positevely associated with wider petals.

Covariances between width of different part: - Sepal length and petal length show positive covariance of 1.27. It means that longer sepals tend to have longer petals. - Sepal width and petal length show negative covariance of -0.12. It means that wider sepals tend very little to narrower petals.

```{r}
cov(iris[, 1:4])
```

## Correlation matrix

The highest correlation in the matrix is between petal length and its width of 0.96. This is extremely positive as petals get longer, they also get wider. The most negative correlation is between sepal width and petal length of -0.43. Since its close to -0.5, it meas that flowers with wider sepals tend to have shorter petals. Interestingly the weakest correlation is between sepal length and its width of -0.12, which means that for that flower part, width and length are not correlated, perhaps slightly negatively.

```{r}
cor(iris[, 1:4])
```

# Correlation Visualization

The correlation matrix can be visualized to highlight the strength and direction of relationships:
Indeed the positive correlations are colored with blue whereas the negative ones are colored towards dark red. The smallest points for the correlation between sepal width and its width are as mentioned the ones with the closest correlation to zero. - The orange points represent the negative correlations described in task 4.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(corrplot)
```

```{r, fig.height=3, fig.width=3}
cor_matrix <- cor(iris[, 1:4])
corrplot(cor_matrix)
```

# Standardization and Covariance Transformation

To link covariance and correlation, we apply a linear transformation based on Equation (1.13).

From Equation 1.13 $Cov(Ax)=ACov(x)A^T$ we need to find a matrix A such that Cov(Ax) = cor(x)

We set as X matrix the observations and compute covariance and correlation matrices.

```{r}
X <- iris[, 1:4]

cov_matrix <- cov(X)   # Σ = Cov(x)
cor_matrix <- cor(X)   # Cor(x)
```

The matrix D is a diagonal matrix that contains the variances of each variable from the covariance matrix on its diagonal. Its square root, $D^{1/2}$ has the standard deviation. Therefore we need to use $A=D^{-1/2}$ to standardize each variable by dividing it by its own standard deviation.

The standardization formula is $z_i=\frac{x_i-\bar{x}_i}{\sigma_i}$

However in matrix form it is $z=D^{-1/2}(\mathbf{x}-\bar{\mathbf{x}})$

```{r}
D <- diag(diag(cov_matrix))

A <- diag(1 / sqrt(diag(cov_matrix)))

```

Here we compute equation 1.13 $Cov(Ax)=ACov(x)A^T$
```{r}

Cov_Ax <- A %*% cov_matrix %*% t(A)

```

We compare if they are the same
```{r}
Cov_Ax
```

```{r}
cor_matrix
```

# Eigenvalue Decomposition of the Covariance Matrix

Every symmetric matrix Σ can be decomposed as $Σ=ΓAΓ^{T}$. Obviously covariance and correlation matrices are symmetrical. A is a diagonal matrix that contains eigenvalues. Γ is the matrix of eigenvectors of Σ. 

Performing eigenvalue decomposιtion

```{r}
eig <- eigen(cov_matrix)
eig
```
Reconstructing Σ from eigen decomposition

```{r}
Gamma <- eig$vectors           
A <- diag(eig$values)          
Sigma_reconstructed <- Gamma %*% A %*% t(Gamma)
Sigma_reconstructed
```
Compare if cov_matrix equals numerically with sigma_reconstructed

```{r}
cov_matrix
```
```{r, eval=FALSE, include=FALSE, echo=FALSE}
## Estimate the correlation matrix of the numeric variables
#Pearson: investigates linear relationsships (default)


cor.pearson <- cor(X)


#Kendall: investigates non-linear relationsships

cor.kendall <- cor(X, method = "kendall")

#Do eigen-decomposition and compare eigenvalues:

eigen.pearson <- eigen(cor.pearson)
eigen.kendall <- eigen(cor.kendall)
plot(eigen.pearson$values,eigen.kendall$values,
     xlab="Eigenvalues from Pearson",ylab="Eigenvalues from Kendall")
abline(c(0,1),col=4,lty=2)
```

# Conclusion

The analysis shows that petal measurements are the most informative variables for distinguishing species, while sepal characteristics exhibit weaker separation. The empirical results also confirm key theoretical properties of covariance transformations and eigenvalue decomposition.
